---
title: "Create cover page"
author: "John Maindonald"
date: "29/06/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{dataIn}
library(tm, quietly=TRUE)
library("SnowballC")
library("RColorBrewer") 
txpages <- pdftools::pdf_text('~/_notes/learnR/learnR.pdf')
nlast <- grep('^Index\ of\ Functions',txpages) - 1 
contpg <- grep('^[[:blank:]]*Contents',txpages)
conts <- paste(c(txpages[contpg],sub('^[[:blank:]]*[[:digit:]]+\n','',txpages[contpg+1])), collapse="")
conts <- strsplit(sub('^[[:blank:]]*Contents\n','',conts), '\n')[[1]]
newchaps <- gsub('[^[:digit:]]+[[:digit:]]?[^[:digit:]]+',':', conts)
chapinfo <- t(sapply(strsplit(newchaps,':'),function(x)as.numeric(x)))
# writeLines(txpages[chapinfo[1,2]:(chapinfo[3,2]-1)], 'texts/12RandEnv.txt')
# writeLines(txpages[chapinfo[3,2]:(chapinfo[4,2]-1)], 'texts/3analysis.txt')
# writeLines(txpages[chapinfo[4,2]:(chapinfo[7,2]-1)], 'texts/456data.txt')
# writeLines(txpages[chapinfo[7,2]:(chapinfo[8,2]-1)], 'texts/7graphs.txt')
# writeLines(txpages[chapinfo[8,2]:(chapinfo[9,2]-1)], 'texts/8reg.txt')
# writeLines(txpages[chapinfo[9,2]:(chapinfo[10,2]-1)], 'texts/9reg.txt')
@

```{r makeTxt}
## In this document, numbering, starts at page 1.
txpages <- textclean::replace_non_ascii(txpages)
ch12 <- paste(txpages[chapinfo[1,2]:(chapinfo[3,2]-1)], collapse=' ')
ch34 <- paste(txpages[chapinfo[3,2]:(chapinfo[5,2]-1)], collapse=' ')
ch56 <- paste(txpages[chapinfo[5,2]:(chapinfo[7,2]-1)], collapse=' ')
ch7 <- paste(txpages[chapinfo[7,2]:(chapinfo[8,2]-1)], collapse=' ')
ch8 <- paste(txpages[chapinfo[8,2]:(chapinfo[9,2]-1)], collapse=' ')
ch9 <- paste(txpages[chapinfo[9,2]:(chapinfo[10,2]-1)], collapse=' ')
processTxt <- function(txts=c(ch12,ch34,ch56,ch7,ch8,ch9)){
  docs <- VCorpus(VectorSource(txts))
  # Convert the text to lower case
  docs <- tm_map(docs, content_transformer(tolower))
  # Remove numbers
  docs <- tm_map(docs, removeNumbers)
  # Remove stopwords for the language 
  docs <- tm_map(docs, removeWords, stopwords("SMART"))
  # Remove punctuations
  docs <- tm_map(docs, removePunctuation)
  # Eliminate extra white spaces
  docs <- tm_map(docs, stripWhitespace)
  # Create term-document matrix
  tdm <- TermDocumentMatrix(docs)
  tdm
}
tdm <- processTxt()
m <- as.matrix(tdm)
getd <- function(mat=m, col=1, add=NA, minfreq=10){
  options(stringsAsFactors = FALSE)
   nam <- rownames(mat)
   freq <- mat[,col]
   v <- order(freq,decreasing=TRUE)
   d <- subset(data.frame(word = nam[v],freq=freq[v]), freq>=10)
   if(!is.na(add))d<-rbind(data.frame(word=add,freq=d$freq[1]+1),d)
   d
}
d1 <- getd(col=1, add='Ch1:2')
d2 <- getd(col=2, add='Ch3:4', minfreq=3); d2[1:2,2] <- d2[1:2,2] - 100
d3 <- getd(col=3, add='ch5:6'); d3[1:3,2] <- d3[1:3,2]-50
d4 <- getd(col=4, add='Ch7')
d5 <- getd(col=5, add='Ch8'); d5[1:3,2] <- d5[1:3,2]-60
d6 <- getd(col=6, add='Ch9')
```

```{wordC}
library(wordcloud2)
pal <- RColorBrewer::brewer.pal(6,'Dark2')[-1]
wordcloud2(d1, size=1.5, color=rep(pal,length.out=nrow(d1)), shape='diamond')  ## ch1:2
wordcloud2(d2, size=1.35, color=rep(pal,length.out=nrow(d2)), shape='diamond')  ## ch34
wordcloud2(d3, size=1.8, color=rep(pal,length.out=nrow(d3)), shape='diamond')  ## ch56
wordcloud2(d4, size=1.75, color=rep(pal,length.out=nrow(d4)), shape='diamond')  ## ch7
wordcloud2(d5, size=1.25, color=rep(pal,length.out=nrow(d5)), shape='diamond')  ## ch8
wordcloud2(d6, size=1.35, color=rep(pal,length.out=nrow(d6)), shape='diamond')  ## ch9
## NB: Use of 'imagemagick' convert
## convert +append b11.png b12.png row1.png
## convert +append b21.png b22.png row2.png
## convert +append b31.png b32.png row3.png
## convert -append a1.png a2.png a3.png a123.png

```


```{gg}
png(file='cover.png',width=2550, height=3300)
library(ggwordcloud)
pal <- adjustcolor(RColorBrewer::brewer.pal(6,'Dark2')[-1],.8)
ggwordcloud2(data=d, size = .5, color = rep(pal, length.out=nrow(d)),
  minRotation = -pi/4, maxRotation = pi/4, shuffle = TRUE,
  rotateRatio = 0.4, shape = "square", ellipticity = 1.3,
  figPath = NA)
ggplot(
  d,
  aes(
    label = word, size = freq,
    color = rep(pal, length.out=nrow(d))),
    angle = 15
  ) +
  geom_text_wordcloud_area(shape='square') +
  scale_size_area(max_size = 12) +
  theme_minimal()
text(1, 8.5, "J H MAINDONALD", adj=0, cex=3)
text(1, 7, "LEARNING AND USING R", adj=0, cex=3)
dev.off()
```